\documentclass[11pt,twoside,titlepage]{report}
\usepackage{moreverb,array,alltt,longtable} %standard
\usepackage{ags-bib,acronyms,ded} %omega
%\usepackage{draftstamp}
\usepackage{keyval}
% here comes the ps/pdftricks stuff, need to run 'pdflatex -shell-escape omdoc'
%\usepackage[shell]{pdftricks}
%\begin{psinputs}
\usepackage{pstricks,pst-node}
%\end{psinputs}
\usepackage[a4paper=true,  bookmarks=true, linkcolor=black,
citecolor=black,urlcolor=black,colorlinks=true,pagecolor=black,
breaklinks=true, bookmarksopen=true]{hyperref}
\usepackage{macros}   %local
\makeindex
\title{{\omdoc}: An Open Markup Format for Mathematical Documents (Version 1.1)}
\author{Michael Kohlhase\\
  Computer Science, Carnegie Mellon University\\
  Pittsburgh, Pa 15213, USA\\
  {\url{http://www.cs.cmu.edu/~kohlhase}}}

\begin{document}
\maketitle
\setcounter{page}{0}\pagenumbering{roman}
\begin{abstract}
  In this report we present a content markup scheme for (collections of)
  mathematical documents including articles, textbooks, interactive books, and
  courses. It can serve as the content language for agent communication of
  mathematical services on a mathematical software bus. We motivate and describe
  the {\omdoc} language and present an {\xml} document type definition for it.
  Furthermore, we discuss applications and tool support.
  
  This document describes version 1.1 of the {\omdoc} format. This version is
  mainly a bug-fix release that has become necessary by the experiments of
  encoding legacy material and theorem prover interfaces in {\omdoc}.  The changes
  are relatively minor, mostly adding optional fields. Version 1.1 of {\omdoc}
  freezes the development so that version 2.0 can be started off.
  
  In contrast to the {\omdoc} format which has not changed much, this report is a
  total re-write, it closes many documentation gaps, clarifies various remaining
  issues. and adds a multitude of new examples.
\end{abstract}
\newpage
\setcounter{tocdepth}{2}\tableofcontents
\newpage\pagenumbering{arabic}
\include{intro}
\include{math-markup}
\include{format}
\include{processing}
%\include{fromlatex}
\include{concl}
\newpage 
\bibliographystyle{ags-alpha} 
\bibliography{omdoc} 
\newpage
{\small\input{\jobname.ind}}
\include{appendix}
\ednotemessage
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
